---
title: "Part II: Inferential Analysis"
subtitle: "for the Statistical Inference Course Project"
author: "Hiten Pragji"
date: "`r format(Sys.Date(), '%d-%m-%Y')`"
output: 
  pdf_document:
    toc: false
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.45in
fontsize: 10pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\large\bfseries}
  - \titleformat*{\subsection}{\normalsize\bfseries}
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries
library(dplyr)
library(ggplot2)
library(datasets)
data(ToothGrowth)
attach(ToothGrowth)
# Set chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

\section{Introduction}

A concern during the Second World War was the provision of vitamin C to soldiers, and in this broad context the effects of ascorbic acid and orange juice were studied in animals. In one such study, 68 guinea pigs, all 28 days old, were given a dietary supplement of vitamin C in one of three doses (0.5, 1, or 2 mg/day) delivered in one of two ways (as ascorbic acid or in orange juice). There were 10 guinea pigs, 5 males and 5 females, in each combination of dose and delivery. After 42 days on the diet, the guinea pigs were sacrificed; incisors were removed and sectioned to obtain measurements on the length of the odontoblasts - cells that are important to tooth development. There could be multiple measurements (odontoblasts) per animal, so the lengths were averaged to give a single value for each animal. The outcome of interest is the average length of the incisor odontoblasts; this was measured in microns. The interest is in how dose and method of delivery influence the outcome.\cite{crampton1947growth}

\section{Aims}

The aim of this analysis is to compare tooth growth by supplement type and dose. To do this, we will import the `ToothGrowth` dataset; the dependent variable (response) is the length of odontoblasts and the independent variables (predictors) are supplement type and dose.

\section{Data description}

```{r explore-data}
# Preview dataset structure
glimpse(ToothGrowth)
str(ToothGrowth)
# Show summary statistics
summary(ToothGrowth)
# View breakdown of dose
ToothGrowth %>%
  dplyr::group_by(dose) %>%
  dplyr::summarise(n = n())
# View breakdown of supplement type
ToothGrowth %>%
  dplyr::group_by(supp) %>%
  dplyr::summarise(n = n())
```

\section{Exploratory data analysis}

```{r eda-plots}
# Plot length x dose
ggplot2::ggplot(
  ToothGrowth,
  aes(x = dose, y = len, colour = supp)
) +
  ggplot2::geom_point()
```

\section{Methods}

\section{Assumptions}

\section{Results}

\section{Discussion}

\section{Conclusion}

\clearpage

\bibliographystyle{IEEEtranN}
\bibliography{inferential}

