---
title: "Part II: Inferential Analysis"
subtitle: "for the Statistical Inference Course Project"
author: "Hiten Pragji"
date: "`r format(Sys.Date(), '%d-%m-%Y')`"
output: 
  pdf_document:
    toc: false
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.75in
fontsize: 10pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\large\bfseries}
  - \titleformat*{\subsection}{\normalsize\bfseries}
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries
library(dplyr)
library(ggplot2)
library(moments)
library(datasets)
data(ToothGrowth)
attach(ToothGrowth)
# Set chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

\section{Introduction}

A concern during the Second World War was the provision of vitamin C to soldiers, and in this broad context the effects of ascorbic acid and orange juice were studied in animals. In one such study, 68 guinea pigs, all 28 days old, were given a dietary supplement of vitamin C in one of three doses (0.5, 1, or 2 mg/day) delivered in one of two ways (as ascorbic acid or in orange juice). There were 10 guinea pigs, 5 males and 5 females, in each combination of dose and delivery. After 42 days on the diet, the guinea pigs were sacrificed; incisors were removed and sectioned to obtain measurements on the length of the odontoblasts - cells that are important to tooth development. There could be multiple measurements (odontoblasts) per animal, so the lengths were averaged to give a single value for each animal. The outcome of interest is the average length of the incisor odontoblasts; this was measured in microns. The interest is in how dose and method of delivery influence the outcome.\cite{crampton1947growth}

\section{Aims}

The aim of this analysis is to compare tooth growth by supplement type and dose. To do this, we will import the `ToothGrowth` dataset; the dependent variable (response) is the length of odontoblasts and the independent variables (predictors) are supplement type and dose.

\section{Data description}

```{r explore-data}
# Preview dataset structure
glimpse(ToothGrowth)
str(ToothGrowth)
# Show summary statistics
summary(ToothGrowth)
# View breakdown of dose
ToothGrowth %>%
  dplyr::group_by(dose) %>%
  dplyr::summarise(n = n())
# View breakdown of supplement type
ToothGrowth %>%
  dplyr::group_by(supp) %>%
  dplyr::summarise(n = n())
```

\section{Exploratory data analysis}

\subsection{Overall distribution of odontoblast length}

```{r eda-plot1}
# Plot basic histogram
ggplot2::ggplot(
  ToothGrowth,
  aes(x = len)
) +
  ggplot2::geom_histogram(
    bins = 20,
    col = "black",
    fill = "lightblue"
  )

moments::skewness(ToothGrowth$len)
moments::kurtosis(ToothGrowth$len)
```

The overall distribution of odontoblast length is near-normal with very slight skewness of -0.15 and the distribution is platykurtic with a kurtosis of 2.02. Although the distribution is not perfectly Gaussian, the observed skewness and kurtosis indicate only minor deviations from normality, and therefore the use of parametric methods such as t-tests remains appropriate.

\subsection{Relationship between dose and odontoblast length}

```{r eda-plot2}
# Plot length x dose
ggplot2::ggplot(
  ToothGrowth,
  aes(x = dose, y = len)
) +
  ggplot2::geom_point()
```

The scatter plot suggests that higher doses of vitamin C, irrespective of supplement type, are associated with increased mean odontoblast length.

\subsection{Effect of supplement type on odontoblast length}

```{r eda-plot3}
ggplot2::ggplot(
  ToothGrowth,
  aes(x = supp, y = len, colour = supp)
) +
  ggplot2::geom_boxplot() +
  ggplot2::facet_grid(~ dose)
```

The boxplots suggest that, for doses of 0.5 mg/day and 1.0 mg/day, supplementation with orange juice was associated with increased mean odontoblast length, however for doses of 2.0 mg/day, the difference in length was not as significant. The natural next step in this analysis is to test, using inference, whether:

\begin{enumerate}[leftmargin=*]
  \item Mean odontoblast length differs by supplement type;
  \item Mean odontoblast length differs by dose;
  \item Dose effect depends on supplement type.
\end{enumerate}

\section{Methods}

\subsection{Hypotheses}

\begin{enumerate}[leftmargin=*]
  \item Effect of supplement type on mean odontoblast length: $H_{0}: \mu_{\text{OJ}} = \mu_{\text{VC}}$, $H_{1}: \mu_{\text{OJ}} \neq \mu_{\text{VC}}$.
  \item Effect of dose on mean odontoblast length: $H_{0}: \mu_{1} = \mu_{2} = \mu_{3}$, $H_{1}: \exists i,j \in {1,2,3}, i \neq j \text{ such that } \mu_{i} \neq \mu_{j}$.
  \item Effect of supplement type on dose response: $H_{0}: \mu_{\text{VC},d} = \mu_{\text{OJ},d}$, $H_{1}: \mu_{\text{VC},d} \neq \mu_{\text{OJ},d}$.
\end{enumerate}

\subsection{Statistical methods}

All hypothesis tests were conducted at a 5$\%$ significance level ($\alpha = 0.5$). All statistical hypothesis tests were two-sided, reflecting non-directional alternative hypotheses.

\subsection{Assumptions}

In this analysis, it is assumed that:

\begin{enumerate}[leftmargin=*]
  \item Mean odontoblast length is approximately normally distributed within each group;
  \item Each length observation is independent;
  \item The variability in mean odontoblast length is different across dose and supplement groups.
\end{enumerate}

\section{Results}

\subsection{Effect of supplement type on odontoblast length}

The following t-test is used to investigate whether there is a significant difference in mean odontoblast length based on the type of supplement (i.e. Hypothesis 1 in Section 5.1):

```{r length-supp-test}
# Create separate dataframes for OJ and VC
oj_len <- ToothGrowth %>%
  dplyr::filter(supp == "OJ") %>%
  dplyr::pull(len)
vc_len <- ToothGrowth %>%
  dplyr::filter(supp == "VC") %>%
  dplyr::pull(len)
# Run two-tailed T-test
t_test_supp <- t.test(
  oj_len,
  vc_len,
  alternative = "two.sided",
  paired = FALSE,
  var.equal = FALSE,
  conf.level = 0.95
)
print(t_test_supp$p.value)
```

Since $p = 0.06 > \alpha$, we fail to reject the null hypothesis; there is insufficient evidence to conclude that the supplement type has a significant impact on the mean length of odontoblasts.

\subsection{Effect of dose on odontoblast length}

The following t-tests are used to investigate whether there is a significant difference in mean odontoblast lengths between different doses (i.e. Hypothesis 2 in Section 5.1):

```{r prep-length-dose-test}
d05 <- ToothGrowth %>%
  dplyr::filter(dose == 0.5) %>%
  dplyr::pull(len)
d10 <- ToothGrowth %>%
  dplyr::filter(dose == 1.0) %>%
  dplyr::pull(len)
d20 <- ToothGrowth %>%
  dplyr::filter(dose == 2.0) %>%
  dplyr::pull(len)
```

```{r length-dose-test}
t_test_dose <- function(d1, d2) {
  test <- t.test(
    d1,
    d2,
    alternative = "two.sided",
    paired = FALSE,
    var.equal = FALSE,
    conf.level = 0.9
  )
  p_value <- test$p.value
  return(p_value)
}
t_test_dose(d05, d10)
t_test_dose(d05, d20)
t_test_dose(d10, d20)
```

In all three cases, since $p < \alpha$, we reject the null hypothesis; this indicates statistically significant differences in mean odontoblast length for all pairwise dose comparisons.

\subsection{Effect of supplement type on dose response}

The following t-tests are used to investigate whether there is a significant difference in mean odontoblast lengths between different doses (i.e. Hypothesis 3 in Section 5.1):

```{r dose-response-test}
doses <- c(0.5, 1.0, 2.0)
t_test_response <- purrr::map(
  doses,
  ~ t.test(
    len ~ supp,
    data = ToothGrowth,
    subset = dose == .x
    )
  )
p_values <- purrr::map_dbl(t_test_response, "p.value")
names(p_values) <- paste0("dose_", doses)
p_values
```

For doses of 0.5 mg/day and 1.0 mg/day, $p < \alpha$, so we reject the null hypothesis; there is sufficient evidence at the 5$\%$ significance level to conclude that mean odontoblast length differs between supplement types. For a dose of 2.0 mg/day however, $p > \alpha$ so we fail to reject the null hypothesis; there is insufficient evidence to conclude a difference in mean odontoblast length between supplement types.

\section{Conclusion}

The results of testing all three hypotheses listed in Section 5.1 are:

\begin{enumerate}
  \item Supplement type does not have a significant effect on mean odontoblast length - as dosage increases, mean odontoblast length increases irrespective of supplememt type;
  \item Dosage of vitamin C does have a significant effect on mean odontoblast length - higher doses of vitamin C contribute to greater tooth length;
  \item For doses of 0.5 mg/day and 1.0 mg/day, there is a significant effect on mean odontoblast length; the same is not true for a dose of 2.0 mg/day.
\end{enumerate}

\bibliographystyle{IEEEtranN}
\bibliography{inferential}
