---
title: "Getting and Cleaning Data"
subtitle: "Johns Hopkins Data Science Project"
author: "Hiten Pragji"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.75in
fontsize: 11pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
---

```{r setup, include=FALSE}
# Load required libraries
library(dplyr)
library(knitr)
library(tinytex)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

\newpage
\section{Introduction}

One of the most exciting areas in all of data science right now is wearable computing. Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained: \href{https://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones}{Human Activity Recognition Using Smartphones}.

Using the data provided, you should create one R script called `run_analysis.R` that does the following:

1. Merges the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement. 
3. Uses descriptive activity names to name the activities in the data set.
4. Appropriately labels the data set with descriptive variable names. 
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

\section{Study design and dataset information}

The dataset is a Human Activity Recognition database built from the recordings of 30 subjects performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors.

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

\section{Analysis}

```{r run-analysis-1}
# Create data subdirectory if it does not already exist
if (!file.exists(here::here("./assignments/getting-cleaning-data/data"))) {
  dir.create(here::here("./assignments/getting-cleaning-data/data"))
}
# Downlaod and extract the zip folder containing the data
zip_file <- here::here("./assignments/getting-cleaning-data/data/har_dataset.zip")
if (!file.exists(zip_file)) {
  data_url <- paste0("https://d396qusza40orc.cloudfront.net/getdata%,
                     2Fprojectfiles%2FUCI%20HAR%20Dataset.zip")
  download.file(data_url, destfile = zip_file, mode = "wb")
  unzip(zip_file, exdir = "./data")
}

# List all files in the data directory
list.files(here::here("./assignments/getting-cleaning-data/data/UCI HAR Dataset/"),
  recursive = TRUE
)
```

The dataset contains the following files relevant to this project:

- 'README.txt'
- 'features_info.txt': Shows information about the variables used on the feature vector.
- 'features.txt': List of all features.
- 'activity_labels.txt': Links the class labels with their activity name.
- 'train/X_train.txt': Training set.
- 'train/y_train.txt': Training labels.
- 'test/X_test.txt': Test set.
- 'test/y_test.txt': Test labels.
- 'test/subject_test.txt': Subject ID for test set.
- 'train/subject_train.txt': Subject ID for training set.

```{r run-analysis-2}
# Read test datasets into data frames
test_x <- read.table(
  here::here(
    "assignments/getting-cleaning-data/data/UCI HAR Dataset/test/X_test.txt"
    )
  )
test_y <- read.table(
  here::here(
    "assignments/getting-cleaning-data/data/UCI HAR Dataset/test/y_test.txt"
    )
  )
test_subject <- read.table(
  here::here(
    "assignments/getting-cleaning-data/data/UCI HAR Dataset/test/subject_test.txt"
    )
  )
# Read training datasets into data frames
train_x <- read.table(
  here::here(
    "assignments/getting-cleaning-data/data/UCI HAR Dataset/train/X_train.txt"
    )
  )
train_y <- read.table(
  here::here(
    "assignments/getting-cleaning-data/data/UCI HAR Dataset/train/y_train.txt"
    )
  )
train_subject <- read.table(
  here::here(
    "./assignments/getting-cleaning-data/data/UCI HAR Dataset/train/subject_train.txt"
    )
  )
# Read features and activity label datasets into data frames
feature_data <- read.table(
  here::here(
    "assignments/getting-cleaning-data/data/UCI HAR Dataset/features.txt"
    )
  )
activity_labels <- read.table(
  here::here(
    "assignments/getting-cleaning-data/data/UCI HAR Dataset/activity_labels.txt"
    )
  )
```

```{r run-analysis-3}
# Rename column headers appropriately
activities <- activity_labels %>%
  dplyr::rename(activity_id = V1, activity_type = V2)
features <- feature_data %>%
  dplyr::rename(activity_id = V1, activity_type = V2)
subject_test <- test_subject %>%
  dplyr::rename(subject_id = V1)
subject_train <- train_subject %>%
  dplyr::rename(subject_id = V1)
test_y_cleaned <- test_y %>%
  purrr::set_names(colnames(activities[1]))
train_y_cleaned <- train_y %>%
  purrr::set_names(colnames(activities[1]))
```

The column names of the test and training datasets (test_x and train_x) should match the variables in the `features` data frame. Currently the variables in these four data frames are labelled $V1, V2, V3, ..., Vn$ so they should be renamed as follows:

```{r run-analysis-4}
# First check that the nrow(features) matches ncol(test_x) and ncol(train_x)
feature_count <- features %>%
  dplyr::summarise(n = n()) %>%
  dplyr::pull(n)
col_count_df <- dplyr::tibble(
  dataset = c("test_x", "test_y", "train_x", "train_y"),
  ncols = c(ncol(test_x), ncol(test_y), ncol(train_x), ncol(train_y))
)

if (feature_count == ncol(test_x) && feature_count == ncol(train_x)) {
  
  test_x_cleaned <- test_x %>%
    purrr::set_names(features$activity_type)

  train_x_cleaned <- train_x %>%
    purrr::set_names(features$activity_type)

} else {
  print("The number of observations in the features data frame does not match \n
  the number of variables in the test/train data frames.")
}
```

After renaming all the variables appropriately, the following merges should be performed:

- Merge test_x_cleaned, test_y_cleaned, and subject_test (1)
- Merge train_x_cleaned, train_y_cleaned, and subject_train (2)
- Merge (1) and (2)

```{r run-analysis-5, message=FALSE}
all_test_merged <- dplyr::bind_cols(
  test_y_cleaned,
  subject_test,
  test_x_cleaned,
  .name_repair = "unique"
)

all_train_merged <- dplyr::bind_cols(
  train_y_cleaned,
  subject_train,
  train_x_cleaned,
  .name_repair = "unique"
)

ml_data <- dplyr::bind_rows(all_test_merged, all_train_merged)
```

The next task is to extract only the measurements on the mean and standard deviation for each measurement. To do this, first retrieve the column names with the words "mean" and "sd" in them using a string detection function, then index the `ml_data` data frame on those columns.

```{r run-analysis-6}
columns <- colnames(ml_data)
mean_sd_colnames <- stringr::str_detect(columns, "activity|subject|mean(?!Freq)|std")
ml_data_filtered <- ml_data[, mean_sd_colnames]
```

The filtered dataset now contains only the measurements on the mean and standard deviation. It also contains the activity ID, but it would be better to include another column showing the name associated with each activity ID.

```{r run-analysis-7}
ml_merged <- ml_data_filtered %>%
  dplyr::left_join(activities,
    by = "activity_id"
  ) %>%
  dplyr::relocate(
    activity_type,
    .after = activity_id
  )
```

This dataset appears to be more complete now. Let's appropriately label the dataset with descriptive variable names using the following mappings:

- t -> time
- f -> frequency
- acc -> accelerometer
- gyro -> gyroscope
- mag -> magnitude
\\ \\
```{r run-analysis-8}
# View all variable names in the dataset
colnames(ml_merged)
# Apply transformations directly to variable names
new_colnames <- purrr::map_chr(
  names(ml_merged), ~ .x %>%
    stringr::str_replace_all("^t", "time") %>%
    stringr::str_replace_all("^f", "frequency") %>%
    stringr::str_replace_all("Acc", "Accelerometer") %>%
    stringr::str_replace_all("Gyro", "Gyroscope") %>%
    stringr::str_replace_all("Mag", "Magnitude")
)
# Assign the new variable names to the data frame
colnames(ml_merged) <- new_colnames
```

From the `ml_merged` dataset, let's compute the average (mean) of each variable for each activity and each subject, then export the data frame to the directory as a text file.

```{r run-analysis-9}
# Create tidy dataset
ml_cleaned <- ml_merged %>%
  dplyr::group_by(activity_type, subject_id) %>%
  dplyr::summarise_all(.funs = mean)
# Write results to file in directory
write.table(ml_cleaned,
            here::here("assignments/getting-cleaning-data/tidyHARdata.txt"),
            row.names = FALSE)
```

The final step is to export all the code chunks into a .R file:

```{r run-analysis-10, echo = TRUE}
# knitr::purl(here::here("assignments/getting-cleaning-data/codebook.Rmd"),
#   output = "assignments/getting-cleaning-data/run_analysis.R"
# )
```

\section{Structure and contents of dataset}

\subsection{Dataset structure}

The `str()` function compactly displays the structure of the `ml_cleaned` data frame, including variable names, types, and the first few values.

```{r structure}
dplyr::glimpse(ml_cleaned)
```

\subsection{Variables}

There are 69 variables in the `ml_cleaned` dataset:

```{r variables}
colnames(ml_cleaned)
```

The dataset includes various measurements from the accelerometer and gyroscope sensors, both in the time and frequency domains. Below is a summary of the key variable groups:

\textbf{\underline{Time Domain Variables}}

These variables represent measurements taken in the time domain, including accelerometer, gyroscope, and jerk signals. Each set of variables corresponds to mean and standard deviation values across different axes (X, Y, Z), or the magnitude of the signals.

Examples:
\begin{itemize}
  \item timeBodyAccelerometer-mean()-X: Mean accelerometer value on the X-axis.
  \item timeBodyAccelerometer-std()-Y: Standard deviation of accelerometer on the Y-axis.
  \item timeBodyGyroscope-mean()-Z: Mean gyroscope value on the Z-axis.
  \item timeBodyGyroscopeJerk-std()-Z: Standard deviation of gyroscope jerk on the Z-axis.
\end{itemize}

\textbf{\underline{Gravity and Jerk Variables}}

These variables capture the gravity component and the jerk (rate of change of acceleration) of the signals from the accelerometer, with both mean and standard deviation calculations for each axis.

Examples:
\begin{itemize}
  \item timeGravityAccelerometer-mean()-X: Mean gravity accelerometer value on the X-axis.
  \item timeBodyAccelerometerJerk-mean()-Y: Mean jerk from the accelerometer on the Y-axis.
\end{itemize}

\textbf{\underline{Magnitude Variables}}

These variables represent the magnitude of the accelerometer, gyroscope, and their respective jerk signals. They include both the mean and standard deviation values.

Examples:
\begin{itemize}
  \item timeBodyAccelerometerMagnitude-mean(): Mean of accelerometer magnitude.
  \item timeBodyGyroscopeJerkMagnitude-std(): Standard deviation of gyroscope jerk magnitude.
\end{itemize}

\textbf{\underline{Frequency Domain Variables}}

These variables represent the measurements in the frequency domain after applying the Fast Fourier Transform (FFT) to the time-domain signals. They include the mean and standard deviation values for accelerometer, gyroscope, and their jerk signals.

Examples:
\begin{itemize}
  \item frequencyBodyAccelerometer-mean()-X: Mean value of accelerometer in the frequency domain on the X-axis.
  \item frequencyBodyGyroscope-std()-Y: Standard deviation of gyroscope in the frequency domain on the Y-axis.
  \item frequencyBodyAccelerometerMagnitude-std(): Standard deviation of accelerometer magnitude in the frequency domain.
\end{itemize}

\textbf{\underline{Frequency Domain Magnitude and Jerk}}

These are the frequency domain equivalents of the magnitude and jerk measurements. They contain mean and standard deviation calculations, reflecting the frequency-domain characteristics of these signals.

Examples:
\begin{itemize}
  \item frequencyBodyBodyAccelerometerJerkMagnitude-mean(): Mean of the accelerometer jerk magnitude in the frequency domain.
  \item frequencyBodyBodyGyroscopeJerkMagnitude-std(): Standard deviation of the gyroscope jerk magnitude in the frequency domain.
\end{itemize}

\section{References}

[1] Reyes-Ortiz, J., Anguita, D., Ghio, A., Oneto, L., & Parra, X. (2013). Human Activity Recognition Using Smartphones [Dataset]. UCI Machine Learning Repository. https://doi.org/10.24432/C54S4K.






































