---
title: "Individual Household Electric Power Consumption"
subtitle: "an exploratory data analyses"
author: "Hiten Pragji"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.75in
fontsize: 11pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
---

```{r setup, include=FALSE}
# Load required libraries
library(data.table)
library(lubridate)
library(ggplot2)
library(dplyr)
library(knitr)
library(tinytex)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

\clearpage
\section{Exploratory data analysis}

Exploratory data analysis is a way of approaching statistics where the aim is to look at data from as many angles as possible in order to observe interesting features. It is much more informal, non-prescriptive, and investigative than its counterpart, confirmatory data analysis (CDA). EDA employs a variety of techniques to:

1. Maximize insight into a data set;
2. Uncover underlying structure;
3. Extract important variables;
4. Detect outliers and anomalies;
5. Test underlying assumptions;
6. Develop parsimonious models;
7. Determine optimal factor settings.

Most EDA techniques are graphical in nature with a few quantitative techniques. The reason for the heavy reliance on graphics is that by its very nature the main the role of EDA is to open-mindedly explore, and graphics gives the analysts unparalleled power to do so.

\section{Project task}

This assignment uses data from the \href{https://archive.ics.uci.edu/}{UC Irvine Machine Learning Repository}, a popular repository for machine learning datasets. We will be using the 'Individual Household Electric Power Consumption' dataset available \href{https://archive.ics.uci.edu/dataset/235/individual+household+electric+power+consumption}{here}. The dataset contains measurements of electric power consumption in one household with a one-minute sampling rate between December 2006 and November 2010 (47 months). Different electrical quantities and some sub-metering values are available.

\section{Data import and transformation}

The dataset contains 2,075,259 rows and 9 columns. Note the following:

- We will only be using data from the dates 2007-02-01 and 2007-02-02. One alternative is to read the data from just those dates rather than reading in the entire dataset and subsetting to those dates.
- You may find it useful to convert the Date and Time variables to Date/Time classes in R using the strptime() and as.Date() functions.
- Note that in this dataset missing values are coded as ?.

Let's import the required data and explore its characteristics.

```{r source-functions}
# Source all required global functions
source("import_data.R")
source("create_day_axis.R")
source("create_legend.R")
source("create_plot1.R")
source("create_plot2.R")
source("create_plot3.R")
source("create_plot4.R")
```

```{r load-data, cache=TRUE}
# Call import_data function to load dataset
power_consumption <- import_data("household_power_consumption.txt")
```

```{r explore-data}
# Check first few rows
head(power_consumption)
# Verify number of variables
ncol(power_consumption)
# Verify number of rows
nrow(power_consumption)
# Check data structure
str(power_consumption)
```

\section{Solutions}

We want to examine how household energy usage varies over a 2-day period in February 2007. Let's perform EDA by creating some plots to describe the variation of energy usage. All graphs to be plotted use base R graphics hence we will use the same.

\underline{\textbf{Plot 1: Histogram of global active power (kW)}}

```{r plot1, fig.width=6, fig.height=6}
# Plot histogram of global active power
create_plot1(power_consumption)
```

```{r plot2}
# Plot line chart of global power consumption over 2 days
create_plot2(power_consumption)
```

```{r plot3}
# Call function to plot energy sub metering plot
create_plot3(power_consumption)
```

The final plot is a 2x2 grid of plots. Let's call each plot x11, x12, x21, x22, starting from the top left corner (x1) and ending at the bottom right corner (x4), from left to right.

```{r plot4, eval=FALSE}
# Call function to 2x2 grid plot
create_plot4(power_consumption)
```






















