---
title: "Hospital Quality Assignment"
author: "Hiten Pragji"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.75in
fontsize: 11pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, include = FALSE}
# Load required libraries
library(tidyr)
library(dplyr)
library(grid)
library(ggplot2)
library(moments)
library(gridExtra)
library(tinytex)
```

\newpage
\section{Introduction}

In this assignment, we study a small subset of data from the Hospital Compare website. On December 1, 2020, the Centers for Medicare & Medicaid Services (CMS) retired Hospital Compare and seven other healthcare quality comparison websites, consolidating their content into a single platform called \href{https://www.medicare.gov/care-compare/}{Care Compare}.

The zip file for this assignment contains three files:

* outcome-of-care-measures.csv: Contains information about 30-day mortality and readmission rates
  + 47 fields
  + Provides each hospital's risk-adjusted 30-day mortality and 30-day readmission category and rate
for heart attacks, heart failure, and pneumonia for over 4,000 hospitals
* hospital-data.csv: Contains information about each hospital
* Hospital_Revised_Flatfiles.pdf: Descriptions of the variables in each file (i.e. the code book).

\section{Task 1}

\textbf{\underline{Plot the 30-day mortality rates for heart attack}}

```{r exercise-1, include = FALSE}
# Read dataset into data frame
outcome <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
# Preview first few rows
head(outcome)
# Preview column names
colnames(outcome)
# Check size of dataset
nrow(outcome)

# View unique values from column: 30-day death rates from heart attack
unique(outcome[, 11])

# Create filtered data frame for numeric data
mortality_rates <- outcome %>%
  dplyr::select(`Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack`) %>%
  dplyr::rename("monthly_mortality_rate" = `Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack`) %>%
  dplyr::mutate(monthly_mortality_rate = as.numeric(monthly_mortality_rate)) %>%
  dplyr::filter(!is.na(monthly_mortality_rate))

# Calculate optimal bin width using Freedman-Diaconis rule
w <- mortality_rates %>%
  dplyr::select(monthly_mortality_rate) %>%
  dplyr::summarise(
    binwidth = (2 * IQR(monthly_mortality_rate, na.rm = TRUE)) / n()^(1 / 3)
  ) %>%
  dplyr::pull(binwidth)

# Create summary statistics to show in ggplot
summary_stats <- mortality_rates %>%
  summarise(
    Mean = mean(monthly_mortality_rate, na.rm = TRUE),
    Median = median(monthly_mortality_rate, na.rm = TRUE),
    SD = sd(monthly_mortality_rate, na.rm = TRUE),
    Skewness = skewness(monthly_mortality_rate, na.rm = TRUE),
    Kurtosis = kurtosis(monthly_mortality_rate, na.rm = TRUE),
    Minimum = min(monthly_mortality_rate, na.rm = TRUE),
    Maximum = max(monthly_mortality_rate, na.rm = TRUE)
  ) %>%
  dplyr::mutate(across(where(is.numeric), ~round(.x, 1)))

summary_stats_transposed <- as.data.frame(t(summary_stats))

colnames(summary_stats_transposed) <- NULL

summary_table <- gridExtra::tableGrob(summary_stats_transposed)

table_grob_resized <- grobWidth(summary_table) * 0.75
table_grob_resized_height <- grobHeight(summary_table) * 0.75

# Create histogram plot using ggplot2
histogram <- ggplot2::ggplot(
  mortality_rates,
  ggplot2::aes(x = monthly_mortality_rate)
) +
  ggplot2::geom_histogram(
    stat = "bin",
    binwidth = w,
    ggplot2::aes(y = after_stat(count)),
    colour = "black",
    fill = "lightgreen"
  ) +
  ggplot2::geom_density(
    ggplot2::aes(y = after_stat(density) * length(mortality_rates$monthly_mortality_rate) * w),
    colour = "red",
    size = 0.25
  ) +
  ggplot2::scale_x_continuous(
    expand = c(0, 0)
  ) +
  ggplot2::scale_y_continuous(
    expand = c(0, 0),
    n.breaks = 10
  ) +
  ggplot2::labs(
    x = "30-day mortality rate",
    y = "Frequency",
    title = "Distribution of 30-day mortality rates for heart attacks"
  ) +
  ggplot2::theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  annotation_custom(
    grob = summary_table,
    xmin = 19, xmax = 21, ymin = 110, ymax = 235
  )
```

```{r print-mortality-plot}
histogram
```

\section{Task 2}

\textbf{\underline{Finding the best hospital in a state}}

```{r best-hospital, echo = TRUE}
best <- function(chosen_state, chosen_outcome) {
  # Import outcome data into data frame
  outcome_data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  # Get unique states
  unique_states <- unique(outcome_data$State)
  
  # Set unique outcomes mapping to column names
  unique_outcomes <- c("heart attack" = "ha_mort_rate", 
                       "heart failure" = "hf_mort_rate", 
                       "pneumonia" = "pne_mort_rate")
  
  # Check that state and outcome are valid
  if (!(chosen_state %in% unique_states)) {
    stop("Invalid state")
  }
  if (!(chosen_outcome %in% names(unique_outcomes))) {
    stop("Invalid outcome")
  }
  
  # Filter dataset based on state and outcome
  filtered_outcome_data <- outcome_data %>%
    # Rename column headers to simplify code
    dplyr::rename(
      "hospital" = Hospital.Name,
      "state" = State,
      "ha_mort_rate" = Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack,
      "hf_mort_rate" = Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure,
      "pne_mort_rate" = Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia
    ) %>%
    # Select the required columns 
    dplyr::select(
      hospital,
      state,
      ha_mort_rate,
      hf_mort_rate,
      pne_mort_rate
      ) %>%
    # Convert mortality rates columns to numeric format
    dplyr::mutate(suppressWarnings(across(3:5, as.numeric))) %>%
    # Keep only complete rows (rows where there are no nulls)
    tidyr::drop_na() %>%
    # Filter state by user-defined input
    dplyr::filter(state == chosen_state) %>%
    # Dynamically select the outcome column based on chosen_outcome
    dplyr::select(
      hospital,
      outcome_rate = dplyr::all_of(unique_outcomes[[chosen_outcome]])
      ) %>%
    # Order by mortality rates (ascending)
    dplyr::arrange(outcome_rate, hospital) %>%
    # Select the first row in the data frame
    dplyr::slice(1) %>%
    # Get the name of the hospital with the lowest mortality rate
    dplyr::pull(hospital)
  
  # (Implicitly) return the filtered data frame
  filtered_outcome_data
}

# Call the function with user-defined parameters
best("AK", "pneumonia")
```

\section{Task 3}

\textbf{\underline{Ranking hospitals by outcome in a state}}

```{r rank-hospital, echo = TRUE}
rankhospital <- function(selected_state, selected_outcome, num = "best") {
  library(dplyr)
  library(tidyr)
  
  # Import outcome data into data frame
  data_outcome <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  # Get unique states
  states_unique <- unique(data_outcome$State)
  
  # Set unique outcomes mapping to column names
  outcomes_unique <- c(
    "heart attack" = "ha_mort_rate", 
    "heart failure" = "hf_mort_rate", 
    "pneumonia" = "pne_mort_rate"
  )
  
  # Check that state and outcome are valid
  if (!(selected_state %in% states_unique)) {
    stop("invalid state")
  }
  if (!(selected_outcome %in% names(outcomes_unique))) {
    stop("invalid outcome")
  }
  
  # Filter dataset based on state and outcome
  outcome_data_filtered <- data_outcome %>%
    # Rename column headers to simplify code
    rename(
      hospital = Hospital.Name,
      state = State,
      ha_mort_rate = Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack,
      hf_mort_rate = Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure,
      pne_mort_rate = Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia
    ) %>%
    # Select the required columns 
    select(hospital, state, ha_mort_rate, hf_mort_rate, pne_mort_rate) %>%
    # Convert mortality rates columns to numeric, suppressing warnings safely
    mutate(across(3:5, ~ suppressWarnings(as.numeric(.)))) %>%
    # Keep only rows where the selected outcome column is not NA
    drop_na(outcomes_unique[[selected_outcome]]) %>%
    # Filter by state
    filter(state == selected_state) %>%
    # Dynamically select the outcome column based on chosen_outcome
    select(hospital, outcome_rate = all_of(outcomes_unique[[selected_outcome]])) %>%
    # Order by mortality rates (ascending) and hospital name
    arrange(outcome_rate, hospital)

  # Convert num to numeric if possible
  num <- suppressWarnings(as.numeric(num))

  # Handle different num values
  if (num == "best") {
    return(outcome_data_filtered$hospital[1])
  } else if (num == "worst") {
    return(outcome_data_filtered$hospital[nrow(outcome_data_filtered)])
  } else if (!is.na(num) && num > 0 && num <= nrow(outcome_data_filtered)) {
    return(outcome_data_filtered$hospital[num])
  } else {
    return(NA)
  }
}

# Call the function with user-defined parameters
rankhospital("TX", "pneumonia", 10)
```

\section{Task 4}

\textbf{\underline{Ranking hospitals in all states}}

We need to write a function named ```rankall``` that takes two arguments: an outcome name (```outcome```) and a hospital ranking (```num```). The function should return a 2-column data frame containing the hospital in each state that has the ranking specified in ```num```. For example, the function call ```rankall("heart attack", "best")``` would return a data frame containing the names of the hospitals that are the best in their respective states for 30-day heart attack death rates.

```{r rank-all, echo = TRUE}
rankall <- function(selected_outcome, num = "best") {
  library(dplyr)
  library(tidyr)
  
  # Import outcome data
  data_outcome <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
  
  # Define valid outcomes and corresponding column names
  outcomes_unique <- c(
    "heart attack" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack",
    "heart failure" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure",
    "pneumonia" = "Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia"
  )
  
  # Check if outcome is valid
  if (!(selected_outcome %in% names(outcomes_unique))) {
    stop("invalid outcome")
  }
  
  # Get the corresponding column name
  outcome_column <- outcomes_unique[[selected_outcome]]
  
  # Process data: select relevant columns, convert to numeric, remove NAs
  ranked_hospitals <- data_outcome %>%
    select(Hospital.Name, State, all_of(outcome_column)) %>%
    mutate(outcome_rate = as.numeric(.data[[outcome_column]])) %>%
    drop_na(outcome_rate) %>%
    group_by(State) %>%
    arrange(outcome_rate, Hospital.Name) %>%
    mutate(rank = row_number()) %>%
    ungroup()
  
  # Select the appropriate rank for each state
  result <- ranked_hospitals %>%
    group_by(State) %>%
    filter(
      (num == "best" & rank == 1) | 
      (num == "worst" & rank == n()) | 
      (is.numeric(num) & num > 0 & num <= n() & rank == num)
    ) %>%
    select(hospital = Hospital.Name, state = State) %>%
    ungroup()
  
  # Ensure every state is represented, adding NA where needed
  all_states <- distinct(data_outcome, State) %>%
    rename(state = State) %>%
    left_join(result, by = "state") %>%
    select(hospital, state)
  
  return(all_states)
}

# Call the function with user-defined parameters
r <- rankall("heart failure", 10)
as.character(subset(r, state == "NV")$hospital)
```
















