---
title: "Data Transformation"
author: "Hiten Pragji"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.75in
fontsize: 11pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
---

```{r setup, include = FALSE}
library(tinytex)
library(knitr)
library(dplyr)
library(ggplot2)
library(nycflights13)

knitr::opts_chunk$set(cache = TRUE)
```

\newpage
\section{Introduction}

In this R Markdown document, I focus on the dplyr package in the tidyverse library. I will illustrate the key ideas around transforming data using data from the nycflights13 package and use the ggplot2 package to help understand and visualise the data.

\section{Exploring the data}

To explore the basic dplyr operations, I will use the `flights` data frame from `nycflights13`. This data frame contains on-time data for 336,776 flights that departed from New York City from JFK, LGA, and EWR airports in 2013. Let's load this into a data frame named `nycflights`:

```{r explore}
nycflights <- nycflights13::flights
nycflights
```

`nycflights` is actually called a tibble, a special type of data frame used by the tidyverse. It differs from a data frame in the way they are printed; tibbles are designed for large datasets, so they show only the first few rows and only the columns that fit on one screen. To view all columns, we can use dplyr::glimpse():

```{r view-all-cols}
dplyr::glimpse(nycflights)
```

In both the tibble view and glimpse view, the variable names are followed by abbreviations that tell you the type of each variable:

- <int>: integer
- <dbl>: double
- <chr>: character
- <dttm>: date-time

\section{dplyr operations}

All dplyr "verbs" have three things in common:

- The first argument is always a data frame
- The subsequent arguments typically describe which columns to operate on
- The output is always a new data frame.

We can perform multiple dplyr operations using the pipe operators `%>%` (magrittr version) or `|>` (base R version). The dplyr verbs are organised into four groups based on what they operate on: rows, columns, groups, and tables.

\subsection{Rows}

\subsubsection{filter()}

`filter()` allows us to keep rows based on values of the columns, specified by some condition(s). Example:

```{r filtering}
nycflights %>%
  dplyr::filter(month == 10)
```

We can also use:

- \> (greater than)
- => (greater than or equal to)
- < (less than)
- <= (less than or equal to)
- == (equal to)
- != (not equal to)
- & or , (and)
- \| (or)

Here is an example of combining multiple conditions in the `filter()` operation:

```{r more-filtering}
nycflights %>%
  dplyr::filter(month %in% c(10, 12),
                sched_dep_time >= 500,
                sched_arr_time <= 900,
                distance < 1000,
                origin == "JFK" | origin == "LGA",
                day != 10)
```

\subsubsection{arrange()}

`arrange()` changes the order of the rows based on the value of the columns. By default, `arrange()` reorders columns from small-to-big. Suppose we want to sort the year, month, day, and departure time in ascending order:

```{r arranging}
nycflights %>%
  arrange(year, month, day, dep_time)
```

To sort a column in descending order, we can use the `desc()` parameter inside the `arrange()` operation:

```{r descending-order}
nycflights %>%
  arrange(year, month, day, desc(dep_time))
```

\subsubsection{distinct()}

`distinct()` finds all the unique rows in a dataset. Let's look at all the unique rows by passing in some variables:

```{r unique-rows}
nycflights %>%
  distinct(origin, dest)
```

To keep the other columns when filtering for unique rows, use the `.keep_all = TRUE` option in `distinct()`.

\subsubsection{Exercises}

1. In a single pipeline for each condition, find all flights that meet the condition:
  - Had an arrival delay of two or more hours
  - Flew to Houston (IAH or HOU)
  - Were operated by United, American, or Delta
  - Departed in summer (July, August, and September)
  - Arrived more than two hours late, but didn't leave late
  - Were delayed by at least an hour, but made up more than 30 minutes in flight.
  
```{r solution-1}
nycflights %>%
  filter(arr_delay >= 120)

nycflights %>%
  filter(dest %in% c("IAH", "HOU"))

nycflights %>%
  filter(carrier %in% c("UA", "AA", "DL"))

nycflights %>%
  filter(month %in% 7:9)

nycflights %>%
  filter(arr_delay > 120, dep_delay == 0)

nycflights %>%
  filter(dep_delay >= 60, dep_delay - arr_delay > 30)
```

2. Sort `flights` to find the flights with the longest departure delays. Find the flights that left earliest in the morning.

```{r solution-2}
nycflights %>%
  arrange(desc(dep_delay))

# Assume "early morning" is between 0000/2400 and 0600 hours
nycflights %>%
  filter(dep_time <= 600 | dep_time == 2400) %>%
  arrange(dep_time)
```

3. Sort `flights` to find the fastest flights. (Hint: Try including a math calculation inside of your function.)

```{r solution-3}
# Speed = distance/time
## Approach 1: Use mutate to create a speed column
nycflights %>%
  mutate(speed = distance/air_time) %>%
  arrange(desc(speed))

## Approach 2: Just use arrange()
nycflights %>%
  arrange(desc(distance/air_time))
```

4. Was there a flight on every day of 2013?

```{r solution-4}
nycflights %>%
  distinct(year, month, day) %>%
  nrow()
```

Yes, there was a flight of every day of 2013.

5. Which flights travelled the farthest distance? Which travelled the least distance)?

```{r solution-5}
# Farthest distance
nycflights %>%
  arrange(desc(distance))

# Least distance
nycflights %>%
  arrange(distance)
```

6. Does it matter what order you used `filter()` and `arrange()` if you're using both? Why/why not? Think about the results and how much work the functions would have to do.

`arrange()` is simply a sorting function to arrange rows based on (a) specific variable(s). The `filter()` function operates on a row-by-row basis using defined conditions, hence there is no difference between the order of operations.

\subsection{Columns}

\subsubsection{mutate()}

`mutate()` allows us to add new columns that are calculated from existing columns. Here is an example of its use:

```{r mutate1}
nycflights %>%
  mutate(
    gain = dep_delay - arr_delay,
    speed = (distance / air_time) * 60
  )
```

To see the new column on the left-hand side, set the `.before` argument to 1:

```{r mutate2}
nycflights %>%
  mutate(
    gain = dep_delay - arr_delay,
    speed = (distance / air_time) * 60,
    .before = 1
  )
```

- `.` is a sign that `.before` is an *argument* to the function, not the name of a third new variable
- Use `.after` to add after a variable
- Variable names can be used instead of positions in both `.before` and `.after`.

We can also control which variables are kept, using the `.keep` argument. For example, to keep only the new (calculated) columns, set `.keep = "used"`:

```{r mutate3}
nycflights %>%
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

\subsubsection{select()}

`select()` allows you to rapidly zoom in on a useful subset using operations based on the names of the variables:

```{r select1}
# Select columns by name:
nycflights %>%
  select(year, month, day)
# Select all columns between year and day (inclusive):
nycflights %>%
  select(year:day)
# Select all columns except those from year to day (inclusive):
nycflights %>%
  select(!year:day)
# Select all columns that are characters:
nycflights %>%
  select(where(is.character))
```

Useful helper functions that can be used within `select()` are:

- `starts_with("abc")`: Matches names that *begin* with "abc"
- `ends_with("xyz")`: Matches names that *end* with "xyz"
- `contains("ijk")`: Matches names that *contain* "ijk"
- `num_range("x", 1:3)`: Matches `x1`, `x2`, `x3`.

We can also rename variables as we select them:

```{r select2}
nycflights %>%
  select(tailnumber = tailnum)
```

\subsubsection{rename()}

`rename()` can be used instead of `select()` to keep all existing variables and just rename a few. The first argument is the new name and the second argument is the old name.

```{r rename-1}
nycflights %>%
  rename(tailnumber = tailnum)
```

If there are many inconsistently named columns, we can automate the renaming using `janitor::clean_names()`.

\subsubsection{relocate()}

`relocate()` is used to move variables around. By default, `relocate()` moves variables to the front:

```{r relocate1}
nycflights %>%
  relocate(time_hour, air_time)
```

We can also specify where to put the variables using the `.before` and `.after` arguments, just like in `mutate()`:

```{r relocate2}
nycflights %>%
  relocate(year:dep_time, .after = time_hour)
nycflights %>%
  relocate(starts_with("arr"), .before = dep_time)
```

\subsubsection{Exercises}

1. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

```{r solution-6}
nycflights %>%
  select(dep_time, sched_dep_time, dep_delay)
```

The relationship is defined by `dep_time = sched_dep_time + dep_delay`.

2. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`.

```{r solution-7}
nycflights %>%
  select(dep_time, dep_delay, arr_time, arr_delay)

nycflights %>%
  select(starts_with("dep"), starts_with("arr"))

nycflights %>%
  select(matches("^dep"), matches("^arr"))
```

3. What happens if you specify the name of the same variable multiple times in a `select()` call?

A variable that is specified multiple times will only appear once:

```{r solution-8}
nycflights %>%
  select(year, year, year, month, day)
```

4. What does the `any_of()` function do? Why might it be helpful in conjunction with this vector: `variables = c("year", "month", "day", "dep_delay", "arr_delay")`?

The `any_of()` function allows you to apply one or more functions across multiple columns. For example:

```{r solution-9}
variables = c("year", "month", "day", "dep_delay", "arr_delay")
nycflights %>%
  select(any_of(variables))
```

5. Does the result of running the following code surprise you? How do the select helpers deal with upper- and lowercase by default? How can you change that default? `flights %>% select(contains("TIME"))`

```{r solution-10}
flights %>%
  select(contains("TIME"))
```

The output is not surprising because the `contains` helper uses base R's grepl() function internally with ignore.case = TRUE, meaning it performs a case-insensitive pattern match by default. You can change the default behaviour like this:

```{r solution-11}
flights %>%
  select(contains("TIME", ignore.case = FALSE))
```

The output gives a tibble with no data because no variables contain the literal string "TIME".

6. Rename `arr_time` to `air_time_min` to indicate units of measurement and move it to the beginning of the data frame.

```{r solution-12}
nycflights %>%
  rename(air_time_min = arr_time) %>%
  relocate(air_time_min, .before = 1)
```

7. Why doesn't the following work, and what does the error mean? `flights %>% select(tailnum) %>% arrange(arr_delay)`

The code does not work because we have attempted to arrange the data frame using a variable (`arr_delay`) that does not exist in the pipeline. The `select()` operation has pulled only the `tailnum` column.

\subsection{Groups}

\subsubsection{group$\_$by()}

Use `group_by()` to divide your dataset into groups meaningful for your analysis. For example, we can group flights by month:

```{r group}
nycflights %>%
  group_by(month)
```

\subsubsection{summarise()}

Use `summarise()` to calculate summary statistics across groups:

```{r summary}
nycflights %>%
  group_by(month) %>%
  summarise(
    average_delay = mean(dep_delay, na.rm = TRUE),
    n = n()
  )
```

Remember to specify `na.rm = TRUE` to ignore all missing values.

\subsubsection{Slicing functions}

There are five handy functions that allow you to extract specific rows within each group:

- `slice_head(n = 1)`: Takes the first row from each group
- `slice_tail(n = 1)`: Takes the last row in each group
- `slice_min(x, n = 1)`: Takes the row with the smallest value of column `x`
- `slice_max(x, n = 1)`: Takes the row with the largest value of column `x`
- `slice_sample(n = 1)`: Takes one random row.

We can vary `n` to select more than one row. For example, the following code finds the flights that are most delayed upon arrival at each destination:

```{r slice}
nycflights %>%
  group_by(dest) %>%
  slice_max(arr_delay, n = 1) %>%
  relocate(dest)
```

\subsubsection{Grouping by multiple variables}

We can create groups using more than one variable:

```{r grouping}
nycflights %>%
  group_by(year, month, day) %>%
  summarise(n = n())
```

Note the console message: `summarise()` has grouped output by `year`, `month`. You can override using the `.groups` argument. What does this mean?

1. `group_by(year, month, day)` groups the data by year, month, and day, meaning any summary you apply will be calculated separately for each day.
2. `summarise(n = n())` counts the number of flights on each day. However, when `summarise()` runs, it automatically removes the last grouping level (`day`) but keeps the higher levels (`year` and `month`).

By default, dplyr removes only the last level of grouping when summarising. This allows for further summaries at the monthly level without having to `group_by(year, month)` again.

You can use the `.groups` argument to change what happens after `summarise()`:

1. If you want to completely remove grouping (ungrouped result):

```{r remove-grouping}
nycflights %>%
  group_by(year, month, day) %>%
  summarise(n = n(), .groups = "drop")
```

Result: No grouping remains.
Consequence: Any further operations will treat it like a normal data frame.

2. If you want to keep all three grouping levels (`year`, `month`, `day`):

```{r keep-grouping}
nycflights %>%
  group_by(year, month, day) %>%
  summarise(n = n(), .groups = "keep")
```

Result: Data remains grouped by `year`, `month`, and `day`.
Consequence: Any further summarisation will still operate at the daily level, unless you explicitly change the grouping.

3. If you want the default behaviour (`drop_last`, keeping `year` and `month`):

```{r default-grouping}
nycflights %>%
  group_by(year, month, day) %>%
  summarise(n = n(), .groups = "drop_last")
```

Result: Data remains grouped by `year` and `month`, but not by `day`.
Consequence: If you summarise again, it will summarise within each month, not across all rows.

In summary:

- The way `summarise()` handles grouping affects what happens when you do further operations on the resulting data frame.
- If you plan to summarise again and need a clean result, use `.groups = "drop"` to remove all grouping.
- If you plan to summarise in steps, keeping grouping (`.groups = "drop_last"` or `.groups = "keep"`) can help avoid re-grouping manually.
- If you get unexpected results (like monthly totals instead of yearly totals), check whether grouping is still active with `group_vars(data)`.

\subsubsection{Ungrouping}

The `ungroup()` function is useful when you want to remove grouping from a data frame after using `group_by()`. This is important because grouping affects how subsequent operations (like `summarise()`, `mutate()`, and `filter()`) behave. Let's see what happens when you summarise an ungrouped data frame:

```{r ungrouping}
nycflights %>%
  group_by(year, month, day) %>%
  ungroup() %>%
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    flights = n()
  )
```

You get a single row back because dplyr treats all the rows in an ungrouped data frame as belonging to one group.

\subsubsection{Exercises}

1. Which carrier has the worst average delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights %>% group_by(carrier, dest) %>% summarize(n())`)

2. Find the flights that are most delayed upon departure from each destination.

3. How do delays vary over the course of the day? Illustrate your answer with a plot.

4. What happens if you supply a negative n to `slice_min()` and friends?

5. Explain what `count()` does in terms of the dplyr verbs you just learned. What does the sort argument to `count()` do?

6. Suppose we have the following tiny data frame:

`df <- tibble(x = 1:5, y = c("a", "b", "a", "a", "b"), z = c("K", "K", "L", "L", "K"))`

a. Write down what you think the output will look like, then check if you were correct, and describe what `group_by()` does.

`df %>% group_by(y)`

b. Write down what you think the output will look like, then check if you were correct, and describe what arrange() does. Also, comment on how it’s different from the group_by() in part (a).

`df %>% arrange(y)`

c. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.

`df %>% group_by(y) %>% summarise(mean_x = mean(x))`

d. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. Then, comment on what the message says.

`df %>% group_by(y, z) %>% summarise(mean_x = mean(x))`

e. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. How is the output different from the one in part (d)?

`df %>% group_by(y, z) %>% summarise(mean_x = mean(x), .groups = "drop")`

f. Write down what you think the outputs will look like, then check if you were correct, and describe what each pipeline does. How are the outputs of the two pipelines different?

`df %>% group_by(y, z) %>% summarise(mean_x = mean(x))`

`df %>% group_by(y, z) %>% mutate(mean_x = mean(x))`



















