---
title: "Loop Functions"
author: "Hiten Pragji"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry: margin=0.75in
fontsize: 11pt
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage[numbers]{natbib}
  - \usepackage{url}
  - \usepackage{caption}
  - \usepackage{notoccite}
  - \usepackage{enumitem}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \usepackage{titlesec}
  - \usepackage{parskip}
  - \usepackage{hyperref}
  - \usepackage{mathtools}
  - \usepackage{bm}
  - \usepackage{dsfont}
  - \usepackage{bigints}
  - \usepackage[nointlimits]{esint}
  - \usepackage{titlesec}
---
\newpage
\section{Introduction}

Loop functions are functions that allow you to execute loop-like behaviour in a compact form. These functions typically have the word 'apply' in them and are particularly convenient when you need to execute a loop on the command line when using R interactively. These functions are some of the more interesting functions of the R language. We will also cover the debugging that comes with R and how to interpret its output to help you find problems in your programs and functions.

Writing for- and while-loops is useful when programming but not particularly easy when working interactively on the command line. There are some functions which implement looping to make life easier:

\begin{itemize}
  \item lapply: Loop over a list and evaluate a function on each element;
  \item sapply: Same as lapply but try to simplify the result;
  \item apply: Apply a function over the margins of an array;
  \item tapply: Apply a function over subsets of a vector;
  \item mapply: Multivariate version of lapply.
\end{itemize}

An auxiliary function **split** is also useful, particularly in conjunction with lapply.

\section{Loop functions - lapply}

\textbf{lapply} takes three arguments:

\begin{enumerate}
  \item A list, x
  \item A function (or the name of a function)
  \item Other arguments via its ... argument
\end{enumerate}

and always returns a list, regardless of the class of the input.

If x is not a list, it will be coerced to a list using as.list. The actual looping is done internally in C code. Let's look at an example:

```{r loopfunc1}
x <- list(a = 1:5, b = rnorm(10))
print(x)
lapply(x, mean)
```

Observe that the list x contains two elements, a and b. The lapply function executes the mean() function on each element, then returns the results as a list. Here's another example:

```{r loopfunc2}
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))
lapply(x, function(elt) elt[, 1])
```

This is an anonymous function for extracting the first column of each matrix.

\section{Loop functions - sapply}

The sapply function will try to simplify the result of lapply if possible:

\begin{itemize}
  \item If the result is a list where every element is length 1, then a vector is returned;
  \item If the result is a list where every element is a vector of the same length (>1), a matrix is returned;
  \item If it cannot determine the object class, a list is returned.
\end{itemize}

Here's an example of the sapply function in action:

```{r loopfunc3}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean) # each element is a list
sapply(x, mean) # sapply simplifies the result as a vector
```

\section{Loop functions - apply}

The apply function is used to evaluate a function (often an anonymous one) over the margins of an array:

\begin{itemize}
  \item It is most often used to apply a function to the rows or columns of a matrix;
  \item It can be used with general arrays, e.g., taking the average of an array of matrices;
  \item It is not really faster than writing a loop, but it works in one line.
\end{itemize}

This is the structure of the apply function:

```{r define_apply}
str(apply)
```

\begin{itemize}
  \item x is an array;
  \item MARGIN is a vector giving the subscripts which the function will be applied over, e.g., for a matrix, 1 indicates rows, 2 indicates columns, 'c(1, 2)' indicates rows and columns;
  \item FUN is a function to be applied;
  \item ... is for other arguments to be passed to FUN.
\end{itemize}

Example: Returning  a vector containing the mean of each row

```{r loopfunc4}
x <- matrix(rnorm(200), 20, 10) # Define normally-distributed random number matrix with 20 rows and 10 columns
print(x)
apply(x, 2, mean) # Apply the mean function to the columns
apply(x, 1, mean) # Apply the mean function to the rows
```

For sums and means of matrix dimensions, we have some shortcuts:

* rowSums = apply(x, 1, sum)
* rowMeans = apply(x, 1, mean)
* colSums = apply(x, 2, sum)
* colMeans = apply(x, 2, mean)

These shortcut functions are much faster but the performance improvement is not noticeable unless we are dealing with a large matrix.

We can also use the apply function in other ways, e.g., on quantiles of the rows of a matrix:

```{r loopfunc5}
x <- matrix(rnorm(200), 20, 10)
apply(x, 1, quantile, probs = c(0.25, 0.75))
```

In this example, the apply function calculates the 25^th^ and 75^th^ percentiles of each row of the matrix.

\section{Loop functions - mapply}

The mapply function is a multivariate version of apply, which applies a function in parallel over a set of arguments. Its structure is defined as:

```{r define_mapply}
str(mapply)
```

where:

* FUN is a function to apply;
* ... contains arguments to apply over;
* MoreArgs is a list of other arguments to FUN;
* SIMPLIFY indicates whether the result should be simplified.

The mapply function allows you to apply a function over two objects, e.g., two lists.

Example: Create a list where each element is a vector of repeated numbers, using the rep function.

```{r loopfunc6}
# Long way
list(rep(1, 4), rep(2, 3), rep(3, 2), rep(4, 1))
# Short way
mapply(rep, 1:4, 4:1)
```

In simple terms, we are repeating the elements of the numeric vector 1:4 according to the frequency 4:1. So, 1 gets repeated 4 times, 2 gets repeated 3 times, etc.

* rep is the function being applied
* 1:4 is the first argument to rep, specifies the values to be repeated
* 4:1 is the second argument to rep, specifies how many times each value is repeated
* mapply applies the rep function element-wise to the arguments.

Now let's look at how to vectorise a function. Consider the following function which generates some random normal noise:

```{r loopfunc7}
noise <- function(n, mean, sd) {
  rnorm(n, mean, sd)
}
noise(5, 1, 2)
```

The function rnorm(n, mean, sd) generates n random values from a normal distribution with the given mean and sd. If we try to pass in a vector of arguments to the noise function, for example:

```{r loopfunc7a}
noise(1:5, 1:5, 2)
```

where:

* mean = 1:5 - a vector of means, c(1, 2, 3, 4, 5)
* sd = 1:5 - a vector of standard deviations, c(1, 2, 3, 4, 5)
* n = 2 - a scalar for the number of samples

R will pair the first elements of mean and sd (1 and 1), then generate 2 random values.

\section{Loop functions - tapply}

The tapply function is used to apply a function over subsets of a vector:

```{r define_tapply}
str(tapply)
```

* x is a vector
* INDEX is a factor or a list of factors (or else they are coerced to factors)
* FUN is a function to be applied
* ... contains other arguments to be passed to FUN
* simplify indicates whether or not we should simplify the result

Consider the following example, in which we simulate the following vector:

```{r loopfunc8}
x <- c(rnorm(10), runif(10), rnorm(10, 1))
print(x)
```

* rnorm(10) generates 10 normally-distributed random numbers
* runif(10) generates 10 random numbers from a uniform distribution on the interval [0, 1]
* rnorm(10, 1) generate 10 normally-distributed random numbers with a mean of 1

This generates a vector of 30 random numbers. Now let's create a factor variable which contains 3 levels:

```{r loopfunc8a}
f <- gl(3, 10)
print(f)
```

We can use tapply to apply a function e.g. mean() over subsets of the vector x, where the subset is defined by f:

```{r loopfunc8b}
tapply(x, f, mean)
```

Here's what happened:

* tapply calculated the mean of x for each level of the factor f
* it splits x into 3 groups based on the levels of f:
  + Group 1: The first 10 elements of x (from rnorm(10))
  + Group 2: The next 10 elements of x (from runif(10))
  + Group 3: The last 10 elements of x (from rnorm(10, 1))
* For each group, tapply computes the mean.

The default setting for the simplify argument is TRUE, however if you specify FALSE, the result is returned as a list:

```{r loopfunc8c}
tapply(x, f, mean, simplify = FALSE)
```

We could also calculate the range of x for each group:

```{r loopfunc9}
tapply(x, f, range)
```

The tapply function is useful because it splits up a vector into smaller components and applies a function (e.g. a summary statistics function) to those components, and then recombines the components.

\section{The split function}

The split function splits a vector or data frame into groups determined by a factor or a list of factors. It is used in data analysis and machine learning, e.g., to create balanced training and testing sets for model evaluation. This is the structure of the split function:

```{r define_split}
str(split)
```

* x is a vector
* f is a factor (or coereced into one) or a list of factors
* drop indicates whether empty factor levels should be dropped

Example: Recall the vector x and factor variable f. Let's run the split function on them:

```{r run_split}
split(x, f)
```

Observe that the split function always returns a list with which we may want to apply one of the looping functions from above. For example, it is common to use split followed by lapply:

```{r split_lapply}
lapply(split(x, f), mean)
```

We could also split a data frame:

```{r split_df}
library(datasets)
head(airquality)
```

Suppose we want to calculate the mean of each column for each month. We can split the data frame by the month column and then use lapply to calculate the mean of specified columns:

```{r split_mean}
s <- split(airquality, airquality$Month)
lapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")]))
```

Since every element of the result of the lapply operation is a list of length 3, we can use sapply instead of lapply to simplify the result to matrix form:

```{r use_sapply}
sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")]))
```

To remove null values from the calculation, we specify na.rm = TRUE as follows:

```{r use_sapply_2}
sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")], na.rm = TRUE))
```

We can also split on multiple levels. Suppose we have more than one factor variable:

```{r split_multlevel}
# Define 10 normally-distributed random variables
x <- rnorm(10)
# Define grouping variables
f1 <- gl(2, 5)
f2 <- gl(5, 2)
f1
f2
# Combine all levels of f1 with all levels of f2: new variable has 10 levels
interaction(f1, f2) # this is the same as list(f1, f2)
```

Interactions can create empty levels:

```{r split_interact}
str(split(x, list(f1, f2)))
```

which we can drop using drop = TRUE:

```{r split_drop}
str(split(x, list(f1, f2), drop = TRUE))
```









